---
title: "Hierarchical ANOSIM of mistletoe visitors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical ANOSIM of mistletoe visitors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6.5, fig.height = 4.5, dpi = 96)
```

This vignette reproduces the community analysis of animal visitors to two
mistletoe species, *Tristerix aphyllus* and *T. verticillatus*, recorded with
camera traps at Las Chinchillas National Reserve, northern Chile, over two
sampling years (doi:10.1016/j.jaridenv.2025.105518). The question is whether
visitor assemblages differ between years and, within each year, between
mistletoe species.

```{r setup}
library(nestedANOSIM)
```

## Data

`mistletoe_visitors` has one row per camera and year. The first three columns
are metadata and the remaining 40 columns hold the number of records of each
visitor taxon. The focal species (*Mimus thenca* and *Sephanoides
sephaniodes*) and unidentified records were already removed, as in the
original analysis.

```{r data}
data(mistletoe_visitors)
dim(mistletoe_visitors)
table(mistletoe_visitors$Year_study, mistletoe_visitors$Mistletoe)

comm <- mistletoe_visitors[, -(1:3)]   # community data only
meta <- mistletoe_visitors[, 1:3]
```

## Hierarchical ANOSIM

`nested_anosim()` computes Bray-Curtis dissimilarities once and runs three
kinds of ANOSIM on them:

1. an overall test on the combined year x mistletoe factor,
2. a test of the main factor (year), and
3. a test of the nested factor (mistletoe species) within each year, here with
   Holm correction for the two tests.

Cameras that recorded none of the retained taxa have an undefined Bray-Curtis
dissimilarity, so they are removed with a warning.

```{r anosim}
res <- nested_anosim(comm   = comm,
                     main   = meta$Year_study,
                     nested = meta$Mistletoe,
                     method = "bray", permutations = 9999,
                     p_adjust = "holm", seed = 123)
```

The pieces of the result can be used directly:

```{r pieces}
res$nested
res$group_summary
```

`res$overall`, `res$main` and `res$nested_objects` are ordinary
`vegan::anosim()` objects, so `summary()` and other vegan methods work on them.

Note that the nested tests are separate ANOSIMs run within each year, not a
variance partition such as a nested PERMANOVA (`vegan::adonis2()`). Report them
as a hierarchical ANOSIM.

## Rank-dissimilarity boxplots

`plot_anosim_box()` is a ggplot2 version of `plot()` for ANOSIM. Each box shows
the ranks of the dissimilarities between groups or within one group. A
positive R means that between-group ranks are higher than within-group ones.

```{r box-main}
plot_anosim_box(res, which = "main", xlab = "Year")
```

```{r box-nested, fig.show = "hold"}
plot_anosim_box(res, which = "First", xlab = "Mistletoe")
plot_anosim_box(res, which = "Second", xlab = "Mistletoe")
```

Many pairs of cameras share no taxa (Bray-Curtis = 1). Those tied pairs get the
same, highest rank, which is why several boxes reach the same top value.

## nMDS ordination

`plot_nmds_hulls()` runs `vegan::metaMDS()` on the stored dissimilarity matrix
and draws one convex hull per group. `groups` refers to the original 87 rows;
the samples removed as empty are dropped automatically.

```{r nmds-all}
grp <- paste(meta$Year_study, meta$Mistletoe, sep = " - ")
out <- plot_nmds_hulls(res, groups = grp, seed = 1)
out$plot
out$stress
```

One *T. aphyllus* camera from the first year (CTa096) sits far from the rest.
It recorded a single visit of one taxon that only three other cameras share,
so its dissimilarities are almost all 1. This is a feature of the data, not of
the ordination.

The function returns the fitted `metaMDS` object and the site scores, so the
same ordination can be redrawn with other groups without refitting:

```{r nmds-mistletoe}
by_species <- plot_nmds_hulls(out$mds, groups = meta$Mistletoe[res$kept],
                              palette = c("#1B9E77", "#D95F02"))
by_species$plot
```

Named palettes from the **ggsci** package (e.g., `palette = "startrek"`) can be
used when it is installed.

## Citation

If you use these data, please cite doi:10.1016/j.jaridenv.2025.105518. To cite
the package, see `citation("nestedANOSIM")`.
